---
title: "Linguistics Lab"
author: "Prof. Tiffany Tang"
date: "`r Sys.Date()`"
format: 
  html:
    code-fold: show
    code-summary: "Show Code"
    code-tools: true
    theme: sandstone
    # theme: cerulean
    embed-resources: true
    lightbox: true
toc: true
number-sections: true
execute: 
  warning: false
  message: false
---

```{r setup, include=FALSE}
here::i_am("notebooks/linguistics_R.qmd")

# source all functions from R/ folder
for (fname in list.files(here::here("R"), pattern = "*.R")) {
  source(here::here(file.path("R", fname)))
}

# path to data
DATA_PATH <- here::here("data")

set.seed(331)
```

# README

In this notebook, we will walk through an unsupervised learning case study based upon a linguistics survey dataset. The dataset contains responses to a survey with 67 questions, where each question has a set of possible responses indicating how a person typically speaks. The goal is to explore this linguistics survey data and to identify dialects, or groups of people who speak similarly, in the United States. Understanding such variation in languages/dialects can provide insights into historical, social, and geographical factors that have molded society over time.

# Load Data

There are two main datasets that we will be working with:

-   `qa_key`: contains survey question and answer key
-   `ling_orig`: contains the linguistics survey response data + the location of the survey respondant

To begin, let's load in the data.

```{r load-data}
# load in data
qa_key <- load_q_and_a_key(DATA_PATH)
ling_orig <- load_ling_data(DATA_PATH)
```

Next, taking a quick look at the data...

::: panel-tabset
#### Q and A Data

```{r}
qa_key
skimr::skim(qa_key)
```

#### Linguistics Data

```{r}
ling_orig
skimr::skim(ling_orig)
```
:::


# Data Explorations

Let's take a look at some of the survey questions and responses.

To get you started, here is some code to plot a map of the survey respondents and their responses to a single question (here, question 50). \[Note: all plotting functions can be found in the `R/plot.R` file.\]

**Play around with the `qid` below to explore different questions.**

```{r}
plt <- plot_ling_map(ling_orig, qa_key, qid = 50)
plotly::ggplotly(plt)
```

If you would like to look at the responses to multiple questions side-by-side, here is some code to help you do this.

**Play around with the `qids` below to explore different questions side-by-side.**

```{r}
#| fig-width: 16
#| fig-height: 12
#| column: page

qids <- c(80, 73, 110, 65)
plt_ls <- list()
for (qid in qids) {
  plt_ls[[as.character(qid)]] <- plot_ling_map(
    ling_orig, qa_key, qid = qid
  )
}
plt <- patchwork::wrap_plots(plt_ls)
plt
```

::: {.callout-note title="More on EDA" collapse="true"}

Are there other ways that you'd like to visualize this data (before proceeding to dimension reduction)?

Two possibly interesting refinements to the above visualizations are:

1.  There are lots of categories with very few responses. What if we collapse these rare responses into an "other" category to simplify the visualization?

```{r}
# collapse responses with <5% of responses into "other" category
# can change `min_prop` to a different value
collapsed_out <- collapse_survey_responses(ling_orig, qa_key, min_prop = 0.05)
collapsed_ling_df <- collapsed_out$ling_data
collapsed_qa_key <- collapsed_out$qa_key
```

```{r}
#| fig-width: 16
#| fig-height: 12
#| column: page

qids <- c(80, 73, 110, 65)
plt_ls <- list()
for (qid in qids) {
  plt_ls[[as.character(qid)]] <- plot_ling_map(
    collapsed_ling_df, collapsed_qa_key, qid = qid
  )
}
plt <- patchwork::wrap_plots(plt_ls)
plt
```

2.  Visualizing each survey data point may introduce some visual bias towards regions with more responses. Here is a visualization showing the most popular response per county.

```{r}
#| fig-width: 16
#| fig-height: 12
#| column: page

qids <- c(80, 73, 110, 65)
plt_ls <- list()
for (qid in qids) {
  plt_ls[[as.character(qid)]] <- plot_ling_map_by_county(
    collapsed_ling_df, collapsed_qa_key, qid = qid
  )
}
plt <- patchwork::wrap_plots(plt_ls)
plt
```

:::

# Dimension Reduction

Instead of examining each survey question one-by-one, we can use dimension reduction techniques to summarize all 67 questions into a few dimensions. This will allow us to more easily visualize the survey data in a lower-dimensional space and potentially identify patterns in the data that are not immediately obvious from just a single survey question.

## Principal Components Analysis

Let's start with arguably the most popular and simple dimension reduction technique - principal components analysis (PCA) using `prcomp()`.

If we apply PCA directly to our raw linguistics survey data:

```{r}
# do PCA
X <- get_X_matrix(ling_orig)
orig_pca_out <- prcomp(X)
orig_scores <- as.data.frame(orig_pca_out$x)
pve1 <- orig_pca_out$sdev^2 / sum(orig_pca_out$sdev^2)
```

We visualize the results of PCA below:

- PCA (uncolored): the traditional PC plot (PC1 versus PC2)
- PCA (colored): the same PC plot but colored by latitude (left) and longitude (right)
- Map: each respondant is plotted on the US map according to their reported latitude and longitude and colored by their PC1 score (left) and PC2 score (right); we view the results at the individual level (top) and aggregated at the county level (bottom)

Note these plots all contain the same information, but are plotted in different ways.

::: {.panel-tabset}

### PCA (uncolored)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_scatter(orig_scores, components = 1:2)
```

### PCA (colored by location)
```{r}
#| echo: false
#| fig-width: 12

pca_plt1 <- plot_dr_scatter(
  orig_scores, components = 1:2, color = get_location(ling_orig, "lat")
) +
  ggplot2::labs(color = "Latitude")
pca_plt2 <- plot_dr_scatter(
  orig_scores, components = 1:2, color = get_location(ling_orig, "long")
) +
  ggplot2::labs(color = "Longitude")
patchwork::wrap_plots(pca_plt1, pca_plt2)
```

### Map
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(orig_scores, ling_orig)
```

### Map (by county)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(orig_scores, ling_orig, by_county = TRUE)
```

:::

Did we find meaningful clusters? Let's take a look at the top PC loadings:

```{r}
as.data.frame(orig_pca_out$rotation) |> 
  tibble::rownames_to_column("question") |> 
  dplyr::mutate(
    dplyr::across(
      tidyselect::starts_with("PC"),
      function(x) {
        data.frame(
          question = question,
          loading = x,
          value = sprintf("%s (%.2f)", question, x)
        ) |> 
          dplyr::arrange(-abs(loading)) |> 
          dplyr::pull(value)
      }
    )
  ) |> 
  dplyr::select(PC1:PC5) |> 
  head()
```


From the PC loadings, we see that the first PC puts a very high weight on Q59. Coloring the points in the PC plot by Q059 shows:

```{r}
#| fig-width: 12

plot_dr_scatter(
  orig_scores, components = 1:2, 
  color = get_answers(ling_orig$Q059, 59, qa_key)
) +
  ggplot2::labs(color = "Q059")
```


Why is the first PC picking up Q59? Because this question has the most number of choices, so it's encoding ranges from 0-21. Since PCA tries to find the direction that maximizes the amount of variance in our data, PCA picks up a direction that is highly aligned with the Q59 responses because of it's large range.

Let's try normalizing the X so that all the questions are on the same scale:

```{r}
scaled_pca_out <- prcomp(X, scale = TRUE)
scaled_scores <- as.data.frame(scaled_pca_out$x)
pve2 <- scaled_pca_out$sdev^2 / sum(scaled_pca_out$sdev^2)
```

::: {.panel-tabset}

### PCA (uncolored)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_scatter(scaled_scores, components = 1:2)
```

### PCA (colored by location)
```{r}
#| echo: false
#| fig-width: 12

pca_plt1 <- plot_dr_scatter(
  scaled_scores, components = 1:2, color = get_location(ling_orig, "lat")
) +
  ggplot2::labs(color = "Latitude")
pca_plt2 <- plot_dr_scatter(
  scaled_scores, components = 1:2, color = get_location(ling_orig, "long")
) +
  ggplot2::labs(color = "Longitude")
patchwork::wrap_plots(pca_plt1, pca_plt2)
```

### Map
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(scaled_scores, ling_orig)
```

### Map (by county)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(scaled_scores, ling_orig, by_county = TRUE)
```

:::

But this still doesn't show much of a pattern with the geography of the US, which we would expect from our EDA. Why? The numeric encoding of categories imposes an ordering and distance between the categories which doesn't really make sense. In other words, if "you" = 1, "y'all" = 2, and "you guys" = 3, then this encoding implies that "you" and "you guys" is twice as different as "you" and "y'all" or "y'all" and "you guys", but this is not necessarily true.

The issue is thus encoding the categorical variables directly to a number. An alternative approach is to one-hot encode these categorical responses.

```{r}
ling_ohe_df <- one_hot_ling_data(ling_orig)
X <- get_X_matrix(ling_ohe_df)
ohe_pca_out <- prcomp(X)
ohe_scores <- as.data.frame(ohe_pca_out$x)
pve3 <- ohe_pca_out$sdev^2 / sum(ohe_pca_out$sdev^2)
```

::: {.panel-tabset}

### PCA (uncolored)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_scatter(ohe_scores, components = 1:2)
```

### PCA (colored by location)
```{r}
#| echo: false
#| fig-width: 12

pca_plt1 <- plot_dr_scatter(
  ohe_scores, components = 1:2, color = get_location(ling_ohe_df, "lat")
) +
  ggplot2::labs(color = "Latitude")
pca_plt2 <- plot_dr_scatter(
  ohe_scores, components = 1:2, color = get_location(ling_ohe_df, "long")
) +
  ggplot2::labs(color = "Longitude")
patchwork::wrap_plots(pca_plt1, pca_plt2)
```

### Map
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(ohe_scores, ling_ohe_df)
```

### Map (by county)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(ohe_scores, ling_ohe_df, by_county = TRUE)
```

:::

What is the dominating the variation explained in the first PC1 now? 

```{r}
as.data.frame(ohe_pca_out$rotation) |> 
  tibble::rownames_to_column("question") |> 
  dplyr::mutate(
    dplyr::across(
      tidyselect::starts_with("PC"),
      function(x) {
        data.frame(
          question = question,
          loading = x,
          value = sprintf("%s (%.2f)", question, x)
        ) |> 
          dplyr::arrange(-abs(loading)) |> 
          dplyr::pull(value)
      }
    )
  ) |> 
  dplyr::select(PC1:PC5) |> 
  head()
```

The first PC is essentially showing us the variation between people who responded and those who did not respond to the questions.

```{r}
#| fig-width: 12

num_nas <- rowSums(get_X_matrix(ling_orig) == 0)
hist(num_nas)

plot_dr_scatter(ohe_scores, components = 1:2, color = num_nas) +
  ggplot2::labs(color = "Number of\nMissing Responses")
```

Let's revise our data cleaning procedure and remove people who left too many questions unanswered, remove the columns corresponding to NA responses, and do PCA. We'll color the points by their latitude/longitude.

```{r}
# remove samples with too few answers
ling_df_cleaned <- remove_samples(ling_orig, min_answers = 50)
# remove_zeros removes the columns corresponding to NA responses
ling_ohe_df <- one_hot_ling_data(ling_df_cleaned, remove_zeros = TRUE)
X <- get_X_matrix(ling_ohe_df)
# do PCA
pca_out <- prcomp(X)
pca_scores <- as.data.frame(pca_out$x)
pve4 <- pca_out$sdev^2 / sum(pca_out$sdev^2)
```

::: {.panel-tabset}

### PCA (uncolored)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_scatter(pca_scores, components = 1:2)
```

### PCA (colored by location)
```{r}
#| echo: false
#| fig-width: 12

pca_plt1 <- plot_dr_scatter(
  pca_scores, components = 1:2, color = get_location(ling_ohe_df, "lat")
) +
  ggplot2::labs(color = "Latitude")
pca_plt2 <- plot_dr_scatter(
  pca_scores, components = 1:2, color = get_location(ling_ohe_df, "long")
) +
  ggplot2::labs(color = "Longitude")
patchwork::wrap_plots(pca_plt1, pca_plt2)
```

### Map
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(pca_scores, ling_ohe_df)
```

### Map (by county)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(pca_scores, ling_ohe_df, by_county = TRUE)
```

:::

The first two PCs now show some association with geographical regions in the US.

Like before, we can look at the PC loadings to find the features that are driving the top PCs

```{r}
as.data.frame(pca_out$rotation) |> 
  tibble::rownames_to_column("question") |> 
  dplyr::mutate(
    dplyr::across(
      tidyselect::starts_with("PC"),
      function(x) {
        data.frame(
          question = question,
          loading = x,
          value = sprintf("%s (%.2f)", question, x)
        ) |> 
          dplyr::arrange(-abs(loading)) |> 
          dplyr::pull(value)
      }
    )
  ) |> 
  dplyr::select(PC1:PC5) |> 
  head()
```

Lastly, to wrap up our PCA investigations, let us take a look at the proportion of variance explained (PVE) by the top PCs. 
For instructive purposes, we will plot the cumulative PVE for each of the four PCA runs we have done so far. 
We do this so as to highlight that **higher PVE is *not* necessarily indicative of a better PCA**. 
Notably, PC1 from the first PCA run (using the raw data) has the highest PVE, but this PC1 is capturing variance that we do not care about. 
In fact, this variance is an unwanted artifact of the way that the data was encoded. 
PVEs across different PCA runs are also not directly comparable since PVE is always measured *relative* to the total amount of variation in the data and the total amount of variation almost always differs across datasets (or even the same dataset, scaled/cleaned/preprocessed differently). 
Consequently, PVE alone should not be used to evaluate the "quality" or "usefulness" of PCA.

```{r}
top_pcs <- 100
pves <- tibble::tibble(
  PC = 1:top_pcs,
  `1-raw` = cumsum(pve1[1:top_pcs]),
  `2-scaled` = cumsum(pve2[1:top_pcs]),
  `3-onehot` = cumsum(pve3[1:top_pcs]),
  `4-cleaned-onehot` = cumsum(pve4[1:top_pcs])
) |> 
  tidyr::pivot_longer(
    cols = -PC,
    names_to = "mode",
    values_to = "PVE"
  )
plt <- ggplot2::ggplot(pves) +
  ggplot2::aes(x = PC, y = PVE, color = mode) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::labs(
    x = "Number of PCs",
    y = "Cumulative PVE",
    color = "Mode"
  ) +
  ggplot2::theme_minimal()
plt
```


## Alternative Dimension Reduction Pipelines

Other things to try:

- What if we had performed PCA at the county level (after aggregating the survey responses by county)?

```{r}
# aggregate_survey_response_by_county() computes the proportion of each response per county
ling_county_df <- aggregate_survey_response_by_county(ling_ohe_df)
X <- get_X_matrix(ling_county_df)

# do PCA
pca_out <- prcomp(X)
pca_scores <- as.data.frame(pca_out$x)
```

::: {.panel-tabset}

### PCA (uncolored)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_scatter(pca_scores, components = 1:2)
```

### PCA (colored by location)
```{r}
#| echo: false
#| fig-width: 12

pca_plt1 <- plot_dr_scatter(
  pca_scores, components = 1:2, color = get_location(ling_county_df, "lat")
) +
  ggplot2::labs(color = "Latitude")
pca_plt2 <- plot_dr_scatter(
  pca_scores, components = 1:2, color = get_location(ling_county_df, "long")
) +
  ggplot2::labs(color = "Longitude")
patchwork::wrap_plots(pca_plt1, pca_plt2)
```

### Map (by county)
```{r}
#| echo: false
#| fig-width: 12

plot_dr_map(pca_scores, ling_county_df, by_county = TRUE)
```

:::

And the corresponding PC loadings:

```{r}
as.data.frame(pca_out$rotation) |> 
  tibble::rownames_to_column("question") |> 
  dplyr::mutate(
    dplyr::across(
      tidyselect::starts_with("PC"),
      function(x) {
        data.frame(
          question = question,
          loading = x,
          value = sprintf("%s (%.2f)", question, x)
        ) |> 
          dplyr::arrange(-abs(loading)) |> 
          dplyr::pull(value)
      }
    )
  ) |> 
  dplyr::select(PC1:PC5) |> 
  head()
```


- What if we had applied tSNE to the raw data, to the top k PCs, to the county-level data, or to other pre-processed data variants? 

Applying tSNE to the individual-level data (after removing people with too many missing responses and after one-hot encoding):
    
```{r}
ling_ohe_df_distinct <- ling_ohe_df |> 
  # need to remove duplicates or else tsne will throw an error
  dplyr::distinct(
    dplyr::across(tidyselect::starts_with("Q")),
    .keep_all = TRUE
  )
X <- get_X_matrix(ling_ohe_df_distinct)
# Note: default in Rtsne performs PCA and keeps 50 components before running tSNE
tsne_out <- Rtsne::Rtsne(X, dims = 2)
tsne_scores <- as.data.frame(tsne_out$Y)

# Scatter (uncolored)
scatter_plt <- plot_dr_scatter(tsne_scores, components = 1:2)

# Scatter (colored by location)
scatter_plt1 <- plot_dr_scatter(tsne_scores, components = 1:2, color = get_location(ling_ohe_df_distinct, "lat")) +
  ggplot2::labs(color = "Latitude", x = "tSNE1", y = "tSNE2")
scatter_plt2 <- plot_dr_scatter(tsne_scores, components = 1:2, color = get_location(ling_ohe_df_distinct, "long")) +
  ggplot2::labs(color = "Longitude", x = "tSNE1", y = "tSNE2")

# Map
map1 <- plot_dr_map(tsne_scores, ling_ohe_df_distinct)
map2 <- plot_dr_map(tsne_scores, ling_ohe_df_distinct, by_county = TRUE)
```

::: {.panel-tabset}

```{r}
#| output: "asis"
#| echo: false
#| fig-width: 12

plt_ls <- list()
plt_ls[["Scatter (uncolored)"]] <- scatter_plt
plt_ls[["Scatter (colored by location)"]] <- patchwork::wrap_plots(scatter_plt1, scatter_plt2)
plt_ls[["Map"]] <- patchwork::wrap_plots(map1, map2, ncol = 1)
for (plt_name in names(plt_ls)) {
  cat(sprintf("\n\n### %s\n\n", plt_name))
  print(plt_ls[[plt_name]])
}
```

:::

Applying tSNE to the county-level data (after removing people with too many missing responses, after one-hot encoding, and after aggregating to get the county-level data):
    
```{r}
ling_county_df_distinct <- ling_county_df |> 
  # need to remove duplicates or else tsne will throw an error
  dplyr::distinct(
    dplyr::across(tidyselect::starts_with("Q")),
    .keep_all = TRUE
  )
X <- get_X_matrix(ling_county_df_distinct)
# Note: default in Rtsne performs PCA and keeps 50 components before running tSNE
tsne_out <- Rtsne::Rtsne(X, dims = 2)
tsne_scores <- as.data.frame(tsne_out$Y)

# Scatter (uncolored)
scatter_plt <- plot_dr_scatter(tsne_scores, components = 1:2)

# Scatter (colored by location)
scatter_plt1 <- plot_dr_scatter(tsne_scores, components = 1:2, color = get_location(ling_county_df_distinct, "lat")) +
  ggplot2::labs(color = "Latitude")
scatter_plt2 <- plot_dr_scatter(tsne_scores, components = 1:2, color = get_location(ling_county_df_distinct, "long")) +
  ggplot2::labs(color = "Longitude")

# Map
map2 <- plot_dr_map(tsne_scores, ling_county_df_distinct, by_county = TRUE)
```

::: {.panel-tabset}

```{r}
#| output: "asis"
#| echo: false
#| fig-width: 12

plt_ls <- list()
plt_ls[["Scatter (uncolored)"]] <- scatter_plt
plt_ls[["Scatter (colored by location)"]] <- patchwork::wrap_plots(scatter_plt1, scatter_plt2)
plt_ls[["Map"]] <- map2
for (plt_name in names(plt_ls)) {
  cat(sprintf("\n\n### %s\n\n", plt_name))
  print(plt_ls[[plt_name]])
}
```

:::
    
- What if we had applied UMAP to the raw data, to the top k PCs, to the county-level data, or to other pre-processed data variants?

Applying UMAP to the individual-level data (after removing people with too many missing responses and after one-hot encoding):

```{r}
ling_ohe_df_distinct <- ling_ohe_df |> 
  # need to remove duplicates or else tsne will throw an error
  dplyr::distinct(
    dplyr::across(tidyselect::starts_with("Q")),
    .keep_all = TRUE
  )
X <- get_X_matrix(ling_ohe_df_distinct)
umap_params <- umap::umap.defaults
umap_params$n_components <- 2
umap_out <- umap::umap(X, config = umap_params)
umap_scores <- as.data.frame(umap_out$layout)

# Scatter (uncolored)
scatter_plt <- plot_dr_scatter(umap_scores, components = 1:2)

# Scatter (colored by location)
scatter_plt1 <- plot_dr_scatter(umap_scores, components = 1:2, color = get_location(ling_ohe_df_distinct, "lat")) +
  ggplot2::labs(color = "Latitude", x = "UMAP1", y = "UMAP2")
scatter_plt2 <- plot_dr_scatter(umap_scores, components = 1:2, color = get_location(ling_ohe_df_distinct, "long")) +
  ggplot2::labs(color = "Longitude", x = "UMAP1", y = "UMAP2")

# Map
map1 <- plot_dr_map(umap_scores, ling_ohe_df_distinct)
map2 <- plot_dr_map(umap_scores, ling_ohe_df_distinct, by_county = TRUE)
```

::: {.panel-tabset}

```{r}
#| output: "asis"
#| echo: false
#| fig-width: 12

plt_ls <- list()
plt_ls[["Scatter (uncolored)"]] <- scatter_plt
plt_ls[["Scatter (colored by location)"]] <- patchwork::wrap_plots(scatter_plt1, scatter_plt2)
plt_ls[["Map"]] <- patchwork::wrap_plots(map1, map2, ncol = 1)
for (plt_name in names(plt_ls)) {
  cat(sprintf("\n\n### %s\n\n", plt_name))
  print(plt_ls[[plt_name]])
}
```

:::

Applying UMAP to the county-level data (after removing people with too many missing responses, after one-hot encoding, and after aggregating to get the county-level data):

```{r}
ling_county_df_distinct <- ling_county_df |> 
  # need to remove duplicates or else umap will throw an error
  dplyr::distinct(
    dplyr::across(tidyselect::starts_with("Q")),
    .keep_all = TRUE
  )
X <- get_X_matrix(ling_county_df_distinct)

umap_params <- umap::umap.defaults
umap_params$n_components <- 2
umap_out <- umap::umap(X, config = umap_params)
umap_scores <- as.data.frame(umap_out$layout)

# Scatter (uncolored)
scatter_plt <- plot_dr_scatter(umap_scores, components = 1:2)

# Scatter (colored by location)
scatter_plt1 <- plot_dr_scatter(umap_scores, components = 1:2, color = get_location(ling_county_df_distinct, "lat")) +
  ggplot2::labs(color = "Latitude")
scatter_plt2 <- plot_dr_scatter(umap_scores, components = 1:2, color = get_location(ling_county_df_distinct, "long")) +
  ggplot2::labs(color = "Longitude")

# Map
map2 <- plot_dr_map(umap_scores, ling_county_df_distinct, by_county = TRUE)
```

::: {.panel-tabset}

```{r}
#| output: "asis"
#| echo: false
#| fig-width: 12

plt_ls <- list()
plt_ls[["Scatter (uncolored)"]] <- scatter_plt
plt_ls[["Scatter (colored by location)"]] <- patchwork::wrap_plots(scatter_plt1, scatter_plt2)
plt_ls[["Map"]] <- map2
for (plt_name in names(plt_ls)) {
  cat(sprintf("\n\n### %s\n\n", plt_name))
  print(plt_ls[[plt_name]])
}
```

:::

- Other things that can be played with: different number of components, different hyperparameters, ...

## Evaluating and Comparing Dimension Reduction Methods

How can we compare the dimension reduction results from different methods? One way is to create a pair plot of the scores from different methods like follows:

```{r}
ling_ohe_df_distinct <- ling_ohe_df |> 
  # need to remove duplicates or else tsne will throw an error
  dplyr::distinct(
    dplyr::across(tidyselect::starts_with("Q")),
    .keep_all = TRUE
  )
X <- get_X_matrix(ling_ohe_df_distinct)

# apply PCA
pca_out <- prcomp(X)
pca_scores <- as.data.frame(pca_out$x)

# apply tSNE
tsne_out <- Rtsne::Rtsne(X, dims = 2)
tsne_scores <- as.data.frame(tsne_out$Y)

# apply UMAP
umap_params <- umap::umap.defaults
umap_params$n_components <- 2
umap_out <- umap::umap(X, config = umap_params)
umap_scores <- as.data.frame(umap_out$layout)

# compare PCA, tSNE, UMAP
scores_tib <- tibble::tibble(
  PC1 = pca_scores$PC1,
  PC2 = pca_scores$PC2,
  tSNE1 = tsne_scores$V1,
  tSNE2 = tsne_scores$V2,
  UMAP1 = umap_scores$V1,
  UMAP2 = umap_scores$V2
)
ggwrappers::plot_pairs(
  scores_tib, columns = 1:6, 
  # color_lower = get_location(ling_ohe_df_distinct, "long"),
  # color_label_lower = "Longitude",
  point_size = 0.01,
  point_alpha = 0.2
)
```

High correlation (either close to 1 or close to -1) indicates that the component scores are similar between the two methods. Note that since the "orientation" of the low-dimensional components is arbitrary, then a correlation close to -1 is equivalent to a correlation close to 1.
