---
title: "checkpoint 2"
output: html_document
date: "2026-04-07"
---

## Data Cleaning 

The OWID CO2 and Greenhouse Gas emissions dataset from Our World in Data is the main dataset used throughout this project. Beginning with this dataset, a group of variables related to emissions, economic output, and energy use was filtered to focus on the main variables that contributes and highlights the objective. Further analysis was conducted and the variables years was restricted to 1990 and later for relevance. The iso_code was used to filter out regional and economic group observations and only keep the individual countries for accuracy and maintaining data. 

Because some key variables had substantial missingness, a broader cleaned dataset for descriptive exploration and a stricter complete-case subset for analyses requiring all core variables was used. 

```{r setup}
library(tidyverse)
odf <- read_csv("../data/owid-co2-data.csv")

odf_clean <- odf %>%
  select(
    country,
    year,
    iso_code,
    co2,
    co2_per_capita,
    gdp,
    population,
    primary_energy_consumption
  ) %>%
  filter(year >= 1990) %>%
  filter(!is.na(iso_code))

colSums(is.na(odf_clean))

nrow(odf_clean)

odf_clean %>%
  summarize(
    complete_rows = sum(complete.cases(co2, gdp, population, primary_energy_consumption)),
    total_rows = n()
  )

odf_clean %>%
  summarize(
    complete_rows = sum(complete.cases(co2, gdp, population, primary_energy_consumption)),
    total_rows = n()
  )
```


A basic graph is shown in the early EDA process. Global CO2 emissions show a strong upward trend from 1990 and onward, reflecting increasing industrial activity and energy demand. Temporary declines are visible around 2008-2009 and 2020, most likely corresponding to global economic disruptions such as the financial crisis of 2008 and 2020 COVID-19 outbreak. Overall, the long-term trend suggests sustained growth in emissions over time.

```{r}
library(ggplot2)
odf_clean %>%
  group_by(year) %>%
  summarize(global_co2 = sum(co2, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = global_co2)) +
  geom_line() +
  labs(
    title = "Global CO2 Emissions Over Time",
    x = "Year",
    y = "Total CO2"
  ) +
  theme_minimal()


odf_model <- odf_clean %>%
  drop_na(co2, gdp, population, primary_energy_consumption)

nrow(odf_clean)
nrow(odf_model)
```




```{r}
odf_model <- odf_clean %>%
  drop_na(co2, gdp, population, primary_energy_consumption)

odf_clean %>%
  summarize(
    total_rows = n(),
    total_countries = n_distinct(country),
    min_year = min(year),
    max_year = max(year)
  )

odf_model %>%
  summarize(
    total_rows = n(),
    total_countries = n_distinct(country),
    min_year = min(year),
    max_year = max(year)
  )


```



## EDA 
To better understand the structure of the dataset and inform modeling choices, global trends, country-level variation, and relationship between economic activity and emissions were examined.

## Global Trends over Time
The aggregate CO2 emissions shows a clear upward trajectory from 1990 onward with rapid growth beginning in the early 2000s. While there is a slight decline around 2020, the overall patterns remains increasing. At a global level, emissions have continued to scale with industrial and economic expansion over time.

## GDP and CO2 Relationship
The scatterplot of GDP vs CO2 emissions shows a strong positive association, though the relationship is not linear. At lower GDP levels, emissions increase gradually while higher GDP values are associated with a wider spread of emissions outcomes. This dispersion across the scatterplot suggests heterogeneity across countried with some high GDP countries maintaining lower emissions, possible due to differences in energy efficiency or policy effects.

## Country Level Trends
Focusing on the top 10 emitting countries, China shows the most dramatic increase, with noticeable peak after 2000 and continues to become a dominant contributor. India also shows steady growth at a lower magntitude. The United States and Germany appear more stable or slightly declining over time, possibly due to changes in energy use or regulation. 

## Total Emissions by Country
When emissions are aggregated across all years, China and the United States account for the largest shares by a substantial margin. The countries following the two contribute significantly less in comparison. These patterns emphasize the inclusion of GDP as a predictor in models and might suggest country-level effects could be important to capture for later usage.

```{r}
library(tidyverse)

theme_proj <- function() {
  theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(size = 10),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

odf_clean %>%
  group_by(year) %>%
  summarize(global_co2 = sum(co2, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = global_co2)) +
  geom_line(color = "salmon2", linewidth = 1.2) +
  labs(
    title = "Global CO2 Emissions Over Time",
    subtitle = "Annual total emissions in the cleaned dataset",
    x = "Year",
    y = "Total CO2 Emissions (tonnes)"
  ) +
  theme_proj()

odf_model %>%
  ggplot(aes(x = gdp, y = co2)) +
  geom_point(color = "orchid1", alpha = 0.5, size = 2) +
  labs(
    title = "GDP vs CO2 Emissions",
    subtitle = "Complete-case observations from 1990 onward",
    x = "GDP",
    y = "CO2 Emissions (tonnes)"
  ) +
  theme_proj()

top_countries <- odf_clean %>%
  group_by(country) %>%
  summarize(total_co2 = sum(co2, na.rm = TRUE)) %>%
  arrange(desc(total_co2)) %>%
  slice(1:10) %>%
  pull(country)

odf_clean %>%
  filter(country %in% top_countries) %>%
  ggplot(aes(x = year, y = co2, color = country)) +
  geom_line(linewidth = 1) +
  labs(
    title = "CO2 Emissions for Top 10 Emitting Countries",
    subtitle = "Country-level trends since 1990",
    x = "Year",
    y = "CO2 Emissions (tonnes)",
    color = "Country"
  ) +
  theme_proj()

odf_clean %>%
  group_by(country) %>%
  summarize(total_co2 = sum(co2, na.rm = TRUE)) %>%
  arrange(desc(total_co2)) %>%
  slice(1:10) %>%
  ggplot(aes(x = reorder(country, total_co2), y = total_co2)) +
  geom_col(fill = "skyblue2") +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Total CO2 Emissions",
    subtitle = "Summed across available years after filtering",
    x = "Country",
    y = "Total CO2 Emissions (tonnes)"
  ) +
  theme_proj()

```



## Extra ggplot for continents
```{r}
region_df <- odf %>%
  select(country, year, co2) %>%
  filter(country %in% c(
    "Africa (GCP)",
    "Asia (GCP)",
    "Europe (GCP)",
    "North America (GCP)",
    "South America (GCP)"
  ))

ggplot(region_df, aes(x = year, y = co2, color = country)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Regional CO2 Emissions Over Time",
    subtitle = "Aggregate regional trends from the OWID dataset",
    x = "Year",
    y = "CO2 Emissions (tonnes)",
    color = "Region"
  ) +
  theme_minimal()

```


## Unsupervised Learning Model

```{r}
pca_data <- odf_model %>%
  select(co2, co2_per_capita, gdp, population, primary_energy_consumption) %>%
  mutate(
    co2 = log1p(co2),
    co2_per_capita = log1p(co2_per_capita),
    gdp = log1p(gdp),
    population = log1p(population),
    primary_energy_consumption = log1p(primary_energy_consumption)
  )

pca_fit <- prcomp(pca_data, center = TRUE, scale. = TRUE)
summary(pca_fit)
pca_fit$rotation
```

## PCA 

The PCA scores did not show separate groups of countries, suggetsed country-level observations vary along continuous dimensions rather than clusters. The loadings plot show that the CO2, GDP, and primary energy consumption contribute similarly to the first PCA while the second PCA mainly distinguishes population from CO2 emissions per capita. These patterns support the interpetation of PC1 as an overall scale dimension and PC2 as a contrast between population size and emissions intensity. 


```{r}
pca_country <- odf_model %>%
  group_by(country) %>%
  summarize(
    co2 = mean(co2, na.rm = TRUE),
    co2_per_capita = mean(co2_per_capita, na.rm = TRUE),
    gdp = mean(gdp, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    primary_energy_consumption = mean(primary_energy_consumption, na.rm = TRUE)
  ) %>%
  mutate(
    co2 = log1p(co2),
    co2_per_capita = log1p(co2_per_capita),
    gdp = log1p(gdp),
    population = log1p(population),
    primary_energy_consumption = log1p(primary_energy_consumption)
  )

pca_fit_country <- prcomp(
  pca_country %>% select(-country),
  center = TRUE,
  scale. = TRUE
)

summary(pca_fit_country)
pca_fit_country$rotation

pca_scores_country <- as_tibble(pca_fit_country$x) %>%
  bind_cols(pca_country %>% select(country))

ggplot(pca_scores_country, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.6, size = 1.8, color = "orchid1") +
  theme_minimal() +
  labs(
    title = "PCA Scores Plot by Country",
    x = "PC1",
    y = "PC2"
  )

loadings_df <- as_tibble(pca_fit$rotation, rownames = "variable")

ggplot(loadings_df, aes(x = PC1, y = PC2, label = variable)) +
  geom_point(size = 3, color = "salmon") +
  geom_text(size = 4, hjust = 0.5, nudge_y = 0.04) +
  theme_minimal() +
  coord_cartesian(xlim = c(0.28, 0.55), ylim = c(-0.85, 0.7), clip = "off") +
  labs(
    title = "PCA Variable Loadings",
    x = "PC1",
    y = "PC2"
  ) +
  theme(
    plot.margin = margin(10, 30, 10, 20)
  )
```



## Predictive Modeling  

To study how the factors contribute to national carbon dioxide emissions, I fit a regression model using GDP, population, and primary consumption as the predictors. Earlier, I did a simple random train and test split with LASSO and a simple linear regression model but after reviewing the structure of the data, this approach was scrapped. 

Randomly assigning observations to training and test sets ignore the time-series structure of the data and can lead to overly optimistic estimates of predictive performance. To address this, time-based resampling was used instead with the caret package. This will ensure that the models trained on earlier years and evaluated on the later years is more appropriate.

I also changed to a log-linear model. EDA showed the CO2 emissions, GDP, population, and primary energy consumption were all rightly-skewed which made it harder to interpret so I had to transform the variables.

To evaluate whether there are redundant predictors, the correlation structure from the log-transformed variables were also examines. Log GDP and log primary energy consumption are still highly correlated. Due to this substantial multicollinearity, I tried to fit a reduced log-linear model that did not include GDP and kept log population and log primary energy consumption as predictors. 

Model performance was evaluated using RMSE, MAE, and r-squared. I also compared both regression models against a baseline model.


```{r}

library(caret)
odf_model <- odf_clean %>%
  drop_na(co2, gdp, population, primary_energy_consumption) %>%
  arrange(year, country)

years_vec <- sort(unique(odf_model$year))

time_slices <- caret::createTimeSlices(
  1:length(years_vec),
  initialWindow = 20,
  horizon = 5,
  fixedWindow = FALSE
)
```

## train/test split

```{r}
last_train_idx <- time_slices$train[[length(time_slices$train)]]
last_test_idx  <- time_slices$test[[length(time_slices$test)]]

train_years <- years_vec[last_train_idx]
test_years  <- years_vec[last_test_idx]

train_data <- odf_model %>%
  filter(year %in% train_years)

test_data <- odf_model %>%
  filter(year %in% test_years)

train_data %>%
  summarize(
    n_rows = n(),
    min_year = min(year),
    max_year = max(year)
  )

test_data %>%
  summarize(
    n_rows = n(),
    min_year = min(year),
    max_year = max(year)
  )

log_train <- train_data %>%
  mutate(
    log_co2 = log1p(co2),
    log_gdp = log1p(gdp),
    log_population = log1p(population),
    log_energy = log1p(primary_energy_consumption)
  )

log_test <- test_data %>%
  mutate(
    log_co2 = log1p(co2),
    log_gdp = log1p(gdp),
    log_population = log1p(population),
    log_energy = log1p(primary_energy_consumption)
  )
```

## Baseline Model on the Log Scale



```{r}
baseline_log_pred <- mean(log_train$log_co2)

baseline_log_rmse <- sqrt(mean((log_test$log_co2 - baseline_log_pred)^2))
baseline_log_mae  <- mean(abs(log_test$log_co2 - baseline_log_pred))
baseline_log_r2   <- 1 - sum((log_test$log_co2 - baseline_log_pred)^2) /
  sum((log_test$log_co2 - mean(log_test$log_co2))^2)

tibble(
  model = "Baseline Mean (log scale)",
  RMSE = baseline_log_rmse,
  MAE = baseline_log_mae,
  R2 = baseline_log_r2
)


lm_log_fit <- lm(
  log_co2 ~ log_gdp + log_population + log_energy,
  data = log_train
)

summary(lm_log_fit)

log_preds <- predict(lm_log_fit, newdata = log_test)

log_rmse <- sqrt(mean((log_test$log_co2 - log_preds)^2))
log_mae  <- mean(abs(log_test$log_co2 - log_preds))
log_r2   <- 1 - sum((log_test$log_co2 - log_preds)^2) /
  sum((log_test$log_co2 - mean(log_test$log_co2))^2)

tibble(
  model = "Full Log-Linear Regression",
  RMSE = log_rmse,
  MAE = log_mae,
  R2 = log_r2
)

log_train %>%
  select(log_gdp, log_population, log_energy) %>%
  cor(use = "complete.obs")
```


## Reduced log-linear regression

The reduced log-linear model performed pretty well like the full log-linear model with just a small increase in RMSE and a decrease in r-squared. I think I will continue to have a reduced model as my final model, too much multicolinearity is going on. 



The coefficient estimates in the reduced model were easier to interpret. Both predictors were positive and highly significant. The coefficient for log primary energy consumption was 0.857, meaning energy use is a dominant predictor in the model. Log population on the other hand, has a smaller positive coefficient of 0.058. Although these coefficients should not be interpreted as isolated causal effects, the reduced model avoids severe overlap among predictors.  

Correlation across the predictors are still present but the reduced-model provided a more interpretable result while trying to preserve strong predictive performance.



```{r}
lm_log_fit_reduced <- lm(
  log_co2 ~ log_population + log_energy,
  data = log_train
)

summary(lm_log_fit_reduced)

reduced_preds <- predict(lm_log_fit_reduced, newdata = log_test)

reduced_rmse <- sqrt(mean((log_test$log_co2 - reduced_preds)^2))
reduced_mae  <- mean(abs(log_test$log_co2 - reduced_preds))
reduced_r2   <- 1 - sum((log_test$log_co2 - reduced_preds)^2) /
  sum((log_test$log_co2 - mean(log_test$log_co2))^2)

tibble(
  model = "Reduced Log-Linear Regression",
  RMSE = reduced_rmse,
  MAE = reduced_mae,
  R2 = reduced_r2
)

results_tbl <- tibble(
  model = c(
    "Baseline Mean (log scale)",
    "Full Log-Linear Regression",
    "Reduced Log-Linear Regression"
  ),
  RMSE = c(
    baseline_log_rmse,
    log_rmse,
    reduced_rmse
  ),
  MAE = c(
    baseline_log_mae,
    log_mae,
    reduced_mae
  ),
  R2 = c(
    baseline_log_r2,
    log_r2,
    reduced_r2
  )
)

results_tbl
```





















